---
title: "NOAA DB"
author: "jbratanov"
date: "Thursday, May 12, 2016"
output: html_document
---

# Data Processing

#### Load NOAA database records
```{r cacheDataRead, cache=TRUE}
# Read in NOAA database records
noaa_DF <- read.csv(bzfile("c:/coursera/ReproducibleData/week4/repdata_data_StormData.csv.bz2"), 
                sep=",",header=TRUE)   

```
#### Identify and validate column data required for study
```{r}
names(noaa_DF)
```
#### Variables used to show the most harmful weather events to population health and economics.
1. Weather events
   + __EVTYPE__ - Event Type

2. Population Health
   + __FATALITIES__ - Human Fatalities
   + __INJURIES__ - Human Injuries

3. Economic Impact
   + __PROPDMG__ - Property Damage
   + __PROPDMGEXP__ - Property damage cost multiplier
   + __CROPDMG__ - Crop Damage
   + __CROPDMGEXP__ - crop damage cost multiplier
   
#### Get columns of data needed in study.
```{r}
storm_DF <- noaa_DF[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP" )]
```
#### Drop records which have no use for the study to tighten up size and usefulness.
```{r}
# drop any record which doesn't have a at least one useful value above zero in key data columns
study_DF <- storm_DF[ which(storm_DF$FATALITIES > 0
                      | storm_DF$INJURIES > 0
                      | storm_DF$PROPDMG > 0
                      | storm_DF$CROPDMG > 0), ]
```
#### Validate Event Types and categorize.
There are events which can be aggregated from similar event types or mis-spellings.
Started with 488 events types, aggregated to 33 event types.
```{r}
# Get event types
#unique(study_DF$EVTYPE)
# convert events to upper case to help clean data
study_DF$EVTYPE <- toupper(study_DF$EVTYPE)

# Manually did interative changes until most events were categorized.

# 1st iteration
study_DF$EVTYPE[grep("*TORNADO*|*FUNNEL*", study_DF$EVTYPE)] <- "TORNADO"
study_DF$EVTYPE[grep("*HURRICANE*|*TYPHOON*", study_DF$EVTYPE)] <- "HURRICANE"
study_DF$EVTYPE[grep("*WIND*|*BURST*|*GUST*|*TURB*", study_DF$EVTYPE)] <- "WIND"
study_DF$EVTYPE[grep("*FIRE*", study_DF$EVTYPE)] <- "FIRE"

study_DF$EVTYPE[grep("*COLD*|*LOW TEMPERATURE*|*WINTRY*|*FREEZE*|*HYPO*", study_DF$EVTYPE)] <- "COLD"
study_DF$EVTYPE[grep("*HEAT*|*HOT*", study_DF$EVTYPE)] <- "HEAT"
study_DF$EVTYPE[grep("*FOG*", study_DF$EVTYPE)] <- "FOG"
study_DF$EVTYPE[grep("*ICE*|*ICY*|*FROST*|*GLAZE*|*SLEET*|*MIXED*", study_DF$EVTYPE)] <- "ICE"
# 2nd iteration
study_DF$EVTYPE[grep("*SNOW*|*BLIZ*", study_DF$EVTYPE)] <- "SNOW"
study_DF$EVTYPE[grep("*HAIL*", study_DF$EVTYPE)] <- "HAIL"
study_DF$EVTYPE[grep("*SURF*|*SEAS*|*MARINE*|*CURRENT*|*TSUNAMI|*COAST*|*BEACH*|*SWELL*|*TIDES*|*WAVE*",
                     study_DF$EVTYPE)] <- "MARINE"
study_DF$EVTYPE[grep("*LAND*|*AVALANCHE*|*SLIDE*", study_DF$EVTYPE)] <- "LANDSLIDE"
# 3rd iteration
study_DF$EVTYPE[grep("*FLOOD*|*STREAM*|*HIGH WATER*|*DROWN*|*DAM*", study_DF$EVTYPE)] <- "FLOOD"
study_DF$EVTYPE[grep("*DRY*|*DROUGHT*", study_DF$EVTYPE)] <- "DRY"
study_DF$EVTYPE[grep("*RAIN*|*WET*", study_DF$EVTYPE)] <- "RAIN"
study_DF$EVTYPE[grep("*LIGHT*", study_DF$EVTYPE)] <- "LIGHTNING"
study_DF$EVTYPE[grep("*DUST*", study_DF$EVTYPE)] <- "DUST"

```
#### Validate cost multiplier value types.
```{r}
# Property cost factors
table(study_DF$PROPDMGEXP)
# Crop cost factors
table(study_DF$CROPDMGEXP)
```
#### Expontential codes to be changed to numeric values
Storm data documentation shows the following validate values:
  * "H" for hundreds
  * "K" for thousands
  * "M" for millions
  * "B" for billions
  * Numerics for exponent values
```{r}
# Change property and crop damage multiplier types "h", k", "m" and "b" to uppercase.
study_DF$PROPDMGEXP <- toupper(study_DF$PROPDMGEXP)
study_DF$CROPDMGEXP <- toupper(study_DF$CROPDMGEXP)

# Change character multiplier to numeric
# if/then/else would be faster here, but got lazy
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "H" ] <- 100
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "K" ] <- 1000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "M" ] <- 1000000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "B" ] <- 1000000000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "-"|study_DF$PROPDMGEXP == ""|study_DF$PROPDMGEXP == "+"] <- 1
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "0" ] <- 1
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "1" ] <- 10
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "2" ] <- 100
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "3" ] <- 1000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "4" ] <- 10000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "5" ] <- 100000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "6" ] <- 1000000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "7" ] <- 10000000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "8" ] <- 100000000
study_DF$PROPDMGEXP[study_DF$PROPDMGEXP == "9" ] <- 1000000000

# Make numeric
study_DF$PROPDMGEXP <- as.numeric(as.character(study_DF$PROPDMGEXP))

# Multiple to get property value
study_DF$PROPVAL <- study_DF$PROPDMG * study_DF$PROPDMGEXP

# Change character multiplier to numeric
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "H" ] <- 100
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "K" ] <- 1000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "M" ] <- 1000000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "B" ] <- 1000000000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "-"|study_DF$CROPDMGEXP == ""|study_DF$CROPDMGEXP == "?"] <- 1
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "0" ] <- 1
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "1" ] <- 10
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "2" ] <- 100
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "3" ] <- 1000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "4" ] <- 10000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "5" ] <- 100000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "6" ] <- 1000000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "7" ] <- 10000000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "8" ] <- 100000000
study_DF$CROPDMGEXP[study_DF$CROPDMGEXP == "9" ] <- 1000000000

# Make numeric
study_DF$CROPDMGEXP <- as.numeric(as.character(study_DF$CROPDMGEXP))

# Multiple to get crop value
study_DF$CROPVAL <- study_DF$CROPDMG * study_DF$CROPDMGEXP

# Property cost factors
table(study_DF$PROPDMGEXP)
# Crop cost factors
table(study_DF$CROPDMGEXP)


```
# Data Processing

## Type of weather events causing death in the USA 1951-2011

```{r}

fatalities <- aggregate (FATALITIES~EVTYPE, study_DF, sum)
fatalities <- fatalities [order(fatalities$FATALITIES, decreasing=TRUE),]
par(mar=c(12, 6, 1, 1))
barplot (height = fatalities$FATALITIES[1:10], names.arg = fatalities$EVTYPE[1:10], las = 2, cex.names= 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Top 10 Event that Caused Fatalities")
title (ylab = "Total number of Fatalities")
```
## Type of weather events causing death in the USA 1951-2011
```{r}
injuries <- aggregate (INJURIES~EVTYPE, study_DF, sum)
injuries <- injuries [order(injuries$INJURIES, decreasing=TRUE),]
par(mar=c(12, 6, 1, 1))
barplot (height = injuries$INJURIES[1:10],
         names.arg = injuries$EVTYPE[1:10], las = 2, cex.names = 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Top 10 Events that Caused Injuries")
title (ylab = "Total number of Injuries", line=4)
```

## Type of Weather Events harmful to Property in the USA 1951-2011

```{r}

property <- aggregate (PROPVAL~EVTYPE, study_DF, sum)
property <- property [order(property$PROPVAL, decreasing=TRUE),]
par(mar=c(12, 6, 1, 1))
barplot (height = property$PROPVAL[1:10]/(1*10^9), names.arg = property$EVTYPE[1:10], las = 2, cex.names= 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Top 10 Event that Caused Property Damage")
title (ylab = "Total amount of dollars in Billions")


```

## Type of Weather Events harmful to Crops in the USA 1951-2011

```{r}

crops <- aggregate (CROPVAL~EVTYPE, study_DF, sum)
crops <- crops [order(crops$CROPVAL, decreasing=TRUE),]
par(mar=c(12, 6, 1, 1))
barplot (height = crops$CROPVAL[1:10]/(1*10^9), names.arg = crops$EVTYPE[1:10], las = 2, cex.names= 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Top 10 Event that Caused Crop Damage")
title (ylab = "Total amount of dollars in Billions")


```





   